---
title: "Albert y Liam First Assignment"
author: "Liam McBride (liam.mcbride01@estudiant.upf.com)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(stats)
```


# Analysis of the Heart Disease Dataset 
Load the data from
[here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). 
The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Perform an EDA on the dataset

### Reading in the Data
```{r}
hdisease <- read.table("./data/heart_disease_dataset.csv", header=T)
hdisease1 <- hdisease %>%
  mutate(
    sex = ifelse(sex == 1, "male", "female"),
    cp = recode(cp, `1` = "typical angina", `2` = "atypical angina", `3` = "non-anginal pain", `4` = "asymptomatic"),
    fbs = ifelse(fbs == 1, T, F ),
    restecg = recode(restecg, `1` = "abnormal", `2` = "probable hypertrophy", `0` = "normal"),
    exang = ifelse(exang == 1, "yes", "no"),
    slope = recode(slope, `1` = "upsloping", `2` = "flat", `3` = "downsloping"),
    thal = recode(thal, `3` = "normal", `6` = "fixed defect", `7` = "reversible defect"),
    num = ifelse(num == 0, "negative diagnosis", "positive diagnosis")
  )
hdisease1[hdisease1 == "?"] <- NA

summary(hdisease1)

ggplot(hdisease1, aes(x = value)) +
  geom_histogram(bins = 10, fill = "blue", color = "black", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  labs(title = "Histograms of Variables", x = "Value", y = "Frequency")
```


## Create visualizations in order to show which variables seem to be more associated with heart disease

```{r}

```


# 2 Difference in mortality rates in hospitalized COVID-19 patients 
Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks

## Reproduce Figure 1 from the publication

```{r}

```


## Reproduce Figure 2 from the publication
but instead of representing the clusters in the annotation, represent the groups (G1 to G4)

```{r}

```

## Improve figure 2 of the publication
Add a second annotation with information of deathm and a third one with information of gender

```{r}

```


# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
